---
title: "Colour volume estimation"
author: "Hugo Gruson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Colour volume estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rgl.useNULL = TRUE)
knitr::knit_hooks$set(webgl = rgl::hook_webgl)
```

# Alphashapes in chromaticity diagrams

```{r}
library(pavo)
data(flowers)
vis_flowers <- vismodel(flowers, visual = "avg.uv")
tcs_flowers <- colspace(vis_flowers)
plot(tcs_flowers)
tetrashape(tcs_flowers, avalue = 0.5)
```

Now let's try to find the smallest $\alpha$ value which does not create any isolated $n-2$ simplex ($\alpha^*$).
We know this value cannot be smaller than the maximum distance to the closest point. So we can begin by identifying this minimal value:

```{r, eval=FALSE}
library(FNN)
distances <- get.knn(tcs_flowers[, c("x", "y", "z")], k = 1)$nn.dist
annd <- max(distances)/2
```

```{r, include=FALSE}
# This is a work around to not depend on FNN
annd <- 0.115302552185513
```

We now know that the most isolated data point lies at a distance `annd`=`r sprintf("%.5f", annd)` of its closest neighbour.

Alphashapes change in a stepwise fashion and many close values will output the same colour volume. We can then start by finding all the threshold values and test only these values:

```{r}
library(alphashape3d)
shapes <- ashape3d(as.matrix(tcs_flowers[, c("x", "y", "z")]), alpha = 0)
avalues <- unname(shapes$tetra[, "rhoT"])
avalues <- sort(avalues)
```

We find `length(avalues)`=`r length(avalues)` threshold values. Fortunately, we don't have to test them all since we know that $\alpha^*$ cannot be smaller than `annd`:

```{r}
avalues <- avalues[avalues >= annd]
```

We compute the alphashapes for all these values

```{r}
shapes <- ashape3d(as.matrix(tcs_flowers[, c("x", "y", "z")]), alpha = avalues)
```

and for each one of them, we test if all points are in a shape:

```{r}
isin <- inashape3d(shapes, indexAlpha = "all", as.matrix(tcs_flowers[, c("x", "y", "z")]))
allin <- vapply(isin, all, logical(1))
```

The smallest value to satisfy this condition is $\alpha^*$:

```{r}
astar_ind <- match(TRUE, allin)
astar <- avalues[astar_ind]
```

We can now visualise the colour volume of the `flowers` dataset computed with the alphashape method, and the $\alpha^*$ value as parameter:

```{r}
plot(tcs_flowers)
tetrashape(tcs_flowers, avalue = astar)
```

# Alphashapes in perceptually uniform spaces

In the previous section, we focused on alpashapes in chromaticity diagrams since it is the most common space where convex hulls (that alphashapes aim at replacing) are used. But it is also possible to use alphashapes in other spaces, such as perceptually uniform spaces.

Let's first build this uniform space and look at the data points in this space:

```{r}
cd_flowers <- coldist(vis_flowers)
xy_flowers <- jnd2xyz(cd_flowers)
plot(xy_flowers)
```

High-level functions to build the alphashape directly in `pavo` have not yet been implemented but you can use the `alphashape3d` directly to compute the alphashape, its volume and display it in a 3D interactive plot.

```{r, webgl=TRUE}
ashape_jnd <- ashape3d(as.matrix(xy_flowers), alpha = 10)
volume_ashape3d(ashape_jnd)
bg3d("white") 
plot(ashape_jnd)
```

