---
title: "Colour volume estimation"
author: "Hugo Gruson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: main.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Colour volume estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.show = "hold"
)
options(rgl.useNULL = TRUE)
```

```{r echo=FALSE, include=FALSE}
library(rgl)
setupKnitr(autoprint = TRUE)
```

Traditionally, colour volume or "colourfulness" of an object has been measured
by computing the volume of the convex hull containing the data points, as
described in [@Stoddard2008]

$\alpha$-shapes are a new tool available in `pavo` to estimate colour volumes, 
while allowing the presence of voids and pockets, which may lead to a more
accurate measurement of "colourfulness" than convex hulls. For more information
on the theoretical background, please report to the related article [@Gruson2020]

# $\alpha$-shapes in chromaticity diagrams

You can plot the colour volume using $\alpha$-shape with the `vol()` function
(for non-interactive plots, `tcsvol()` otherwise) by specifying 
`type = "alpha"`. By default, this will use the $\alpha^*$ value defined in 
@Gruson2020.

```{r}
library(pavo)
data(flowers)
vis_flowers <- vismodel(flowers, visual = "avg.uv")
tcs_flowers <- colspace(vis_flowers)
plot(tcs_flowers)
vol(tcs_flowers, type = "alpha")
```

To get the colour volume value, you can use the `summary.colspace()` function. 
The colour volume computed by the convex hull is contained in the `c.vol` column
and the colour volume computed by the $\alpha$-shape (with parameter $\alpha^*$)
is contained in the `a.vol` column:

```{r}
summary(tcs_flowers)
```

Alternatively, you can set the $\alpha$ parameter to the value of your choice via the `avalue` 
argument:

```{r}
plot(tcs_flowers)
vol(tcs_flowers, type = "alpha", avalue = 0.5)
```

# $\alpha$-shapes in perceptually uniform spaces

In the previous section, we focused on $\alpha$-shapes in chromaticity diagrams since
it is the most common space where convex hulls (that $\alpha$-shapes aim at
replacing) are used. But it is also possible to use $\alpha$-shapes in other spaces,
such as perceptually uniform spaces.

Let's first build this uniform space and look at the data points in this space:

```{r}
cd_flowers <- coldist(vis_flowers)
xy_flowers <- jnd2xyz(cd_flowers)
plot(xy_flowers)
```

High-level functions to build the $\alpha$-shape directly in `pavo` have not yet
been implemented but you can use the `alphashape3d` package directly to compute
the $\alpha$-shapes, its volume and display it in a 3D interactive plot.

```{r}
library(alphashape3d)
ashape_jnd <- ashape3d(as.matrix(xy_flowers), alpha = 10)
volume_ashape3d(ashape_jnd)
```

```{r, echo = FALSE}
rgl::bg3d("white") 
```

```{r}
plot(ashape_jnd)
```

# Colour volume overlap with $\alpha$-shapes

$\alpha$-shapes can also be used to measure the colour similarity of two objects, 
by computing the colour volume overlap. This is done in `pavo` with the 
`overlap3d()` function. For example, let's compare the colour volume of the 
crown and the breast of stripe-tailed yellow finches (*Sicalis citrina*):


```{r}
data(sicalis)
tcs.sicalis.C <- subset(colspace(vismodel(sicalis)), "C")
tcs.sicalis.B <- subset(colspace(vismodel(sicalis)), "B")
```

```{r, echo = FALSE}
rgl::bg3d()
```

```{r}
voloverlap(tcs.sicalis.C, tcs.sicalis.B, type = "alpha", plot = TRUE)
```

<!-- # Exploring other values of $\alpha$ -->

<!-- Depending on your biological question, it is possible you might want a different -->
<!-- value of $\alpha$ than the $\alpha^*$ suggested earlier.  -->

<!-- For example, you may want all your colour volumes to only have one component, -->
<!-- instead of several multiple smaller volumes. Here is a quick demo of how you can -->
<!-- do this. -->

<!-- First, let's create a dummy dataset with two distinct clouds of points: -->

<!-- ```{r} -->
<!-- set.seed("20200126") -->
<!-- dvol = data.frame( -->
<!--   "x" = c(rnorm(n = 10, mean = -0.7/sqrt(2)+.3, sd = 0.05), -->
<!--           rnorm(n = 10, mean =  0.7/sqrt(2)-.3, sd = 0.05)), -->
<!--   "y" = c(rnorm(n = 10, mean = -0.25*sqrt(2/3)+.3, sd = 0.05), -->
<!--           rnorm(n = 10, mean = -0.25*sqrt(2/3)+.3  , sd = 0.05)), -->
<!--   "z" = c(rnorm(n = 10, mean = -0.5*sqrt(2/3)+.3, sd = 0.05), -->
<!--           rnorm(n = 10, mean = -0.5*sqrt(2/3)+.3  , sd = 0.05)) -->
<!-- ) -->
<!-- ``` -->

<!-- With the approach highlighted before, this will result in a colour volume with -->
<!-- two distinct components: -->

<!-- ```{r} -->
<!-- tetraplot(dvol) -->
<!-- vol(dvol, type= "alpha") -->
<!-- ``` -->

<!-- To prevent this, we can add an additional test with the following code: -->

<!-- ```{r} -->
<!-- library(alphashape3d) -->
<!-- tetras <- ashape3d(as.matrix(dvol), alpha = 0)$tetra -->

<!-- tetras <- tetras[order(tetras[, "rhoT"]), ] -->

<!-- i <- 0 -->
<!-- all_in <- FALSE -->

<!-- while (!all_in && i < nrow(tetras)) { -->
<!--   i <- i+1 -->

<!--   points_in <- unique(c(tetras[seq_len(i), 1:4])) -->

<!--   all_in <- all(seq_len(nrow(dvol)) %in% points_in) -->
<!-- } -->

<!-- a_1 <- tetras[i, "rhoT"] -->
<!-- ``` -->


<!-- ```{r} -->
<!-- tetraplot(dvol) -->
<!-- vol(dvol, type= "alpha", avalue = tetras[i+20, "rhoT"]) -->
<!-- ``` -->
