---
title: "Colour volume estimation"
author: "Hugo Gruson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Colour volume estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rgl.useNULL = TRUE)
knitr::knit_hooks$set(webgl = rgl::hook_webgl)
```

Alphashapes are a new tool available in `pavo` to estimated colour volumes, 
while allowing the presence of voids and pockets, which may lead to a more
accurate measurement of "colourfulness" than convex hulls.

# Alphashapes in chromaticity diagrams

You can plot the colour volume using $\alpha$-shape with the `tetrashape()` 
function (for non-interactive plots, `tcsshape()` otherwise), with a 
user-defined value for $\alpha$, set by the `avalue` argument:

```{r}
library(pavo)
data(flowers)
vis_flowers <- vismodel(flowers, visual = "avg.uv")
tcs_flowers <- colspace(vis_flowers)
plot(tcs_flowers)
tetrashape(tcs_flowers, avalue = 0.5)
```

Now let's try to find the smallest $\alpha$ value which does not create any
isolated $n-2$ simplex ($\alpha^*$). We know this value cannot be smaller than
the maximum distance to the closest point. So we can begin by identifying this
minimal value:

```{r, eval=FALSE}
library(FNN)
distances <- get.knn(tcs_flowers[, c("x", "y", "z")], k = 1)$nn.dist
annd <- max(distances)/2
```

```{r, include=FALSE}
# This is a workaround to not depend on FNN
annd <- 0.115302552185513
```

We now know that the most isolated data point lies at a distance `annd`=`r
sprintf("%.5f", annd)` of its closest neighbour.

Alphashapes change in a stepwise fashion and many close values will output the
same colour volume. We can then start by finding all the threshold values and
test only these values:

```{r}
library(alphashape3d)
shapes <- ashape3d(as.matrix(tcs_flowers[, c("x", "y", "z")]), alpha = 0)
avalues <- unname(shapes$tetra[, "rhoT"])
avalues <- sort(avalues)
```

We find `length(avalues)`=`r length(avalues)` threshold values. Fortunately, we
don't have to test them all since we know that $\alpha^*$ cannot be smaller than
`annd`:

```{r}
avalues <- avalues[avalues >= annd]
```

We compute the alphashapes for all these values

```{r}
shapes <- ashape3d(as.matrix(tcs_flowers[, c("x", "y", "z")]), alpha = avalues)
```

and for each one of them, we test if all points are in a shape:

```{r}
isin <- inashape3d(shapes, indexAlpha = "all", as.matrix(tcs_flowers[, c("x", "y", "z")]))
allin <- vapply(isin, all, logical(1))
```

The smallest value to satisfy this condition is $\alpha^*$:

```{r}
astar_ind <- match(TRUE, allin)
astar <- avalues[astar_ind]
```

We can now visualise the colour volume of the `flowers` dataset computed with
the alphashape method, and the $\alpha^*$ value as parameter:

```{r}
plot(tcs_flowers)
tetrashape(tcs_flowers, avalue = astar)
```

# Alphashapes in perceptually uniform spaces

In the previous section, we focused on alpashapes in chromaticity diagrams since
it is the most common space where convex hulls (that alphashapes aim at
replacing) are used. But it is also possible to use alphashapes in other spaces,
such as perceptually uniform spaces.

Let's first build this uniform space and look at the data points in this space:

```{r}
cd_flowers <- coldist(vis_flowers)
xy_flowers <- jnd2xyz(cd_flowers)
plot(xy_flowers)
```

High-level functions to build the alphashape directly in `pavo` have not yet
been implemented but you can use the `alphashape3d` directly to compute the
alphashape, its volume and display it in a 3D interactive plot.

```{r}
ashape_jnd <- ashape3d(as.matrix(xy_flowers), alpha = 10)
volume_ashape3d(ashape_jnd)
```

```{r, echo = FALSE}
rgl::bg3d("white") 
```

```{r, webgl=TRUE}
plot(ashape_jnd)
```

# Colour volume overlap with alphashapes

Alphashapes can also be used to measure the colour similarity of two objects, by
computing the colour volume overlap. This is done in `pavo` with the 
`overlap3d()` function. For example, let's compare the colour volume of the 
crown and the breast of stripe-tailed yellow finches (*Sicalis citrina*):


```{r}
data(sicalis)
tcs.sicalis.C <- subset(colspace(vismodel(sicalis)), "C")
tcs.sicalis.B <- subset(colspace(vismodel(sicalis)), "B")
```

```{r, echo = FALSE}
rgl::bg3d()
```

```{r, webgl=TRUE}
overlap3d(tcs.sicalis.C, tcs.sicalis.B, avalue = 0.2, plot = TRUE)
```

<!--
This doesn't work right now because it returns an a that is too small

# Exploring other values of alpha

Depending on your biological question, it is possible you might want a different
value of $\alpha$ than the $\alpha^*$ suggested earlier. 

For example, you may want all your colour volume to only have one component,
instead of several multiple smaller volumes. Here is a quick demo of how you can
do this.


First, let's create a dummy dataset with two distinct clouds of points:

```{r, include = FALSE}
set.seed("20200126")
```

```{r}
dvol = data.frame(
  "x" = c(rnorm(n = 10, mean = -0.7/sqrt(2)+.3, sd = 0.05),
          rnorm(n = 10, mean =  0.7/sqrt(2)-.3, sd = 0.05)),
  "y" = c(rnorm(n = 10, mean = -0.25*sqrt(2/3)+.3, sd = 0.05),
          rnorm(n = 10, mean = -0.25*sqrt(2/3)+.3  , sd = 0.05)),
  "z" = c(rnorm(n = 10, mean = -0.5*sqrt(2/3)+.3, sd = 0.05),
          rnorm(n = 10, mean = -0.5*sqrt(2/3)+.3  , sd = 0.05))
)
```

With the approach highlighted before, this will result in a colour volume with
two distinct components:

```{r, eval = F}
distances <- get.knn(dvol, k = 1)$nn.dist
annd <- max(distances)/2
```

```{r, include=FALSE}
# This is a workaround to not depend on FNN
annd <- 0.03829851
```

```{r}
shapes <- ashape3d(as.matrix(dvol), alpha = 0)
avalues <- unname(shapes$tetra[, "rhoT"])
avalues <- sort(avalues)
avalues <- avalues[avalues >= annd]
shapes <- ashape3d(as.matrix(dvol), alpha = avalues)
isin <- inashape3d(shapes, indexAlpha = "all", as.matrix(dvol))
allin <- vapply(isin, all, logical(1))
astar_ind <- match(TRUE, allin)
astar <- avalues[astar_ind]
tetraplot(dvol)
tetrashape(dvol, avalue = astar)
```

To prevent this, we can add an additional test that uses 
`components_ashape3d()`. This function tells you in which component is each data
point:

```{r}
comp <- components_ashape3d(shapes, indexAlpha = "all")
```

So, to ensure we only have one component, we can do:

```{r}
nb_comp <- sapply(comp, function(x) length(unique(x)) == 1)
```

We find the first `TRUE` value with `match()`:

```{r}
a_ind <- match(TRUE, nb_comp)
a <- avalues[max(a_ind, astar_ind)]
```

```{r}
tetraplot(as.data.frame(dvol))
tetrashape(dvol, avalue = a)
```

-->
